Staged Abstract Interpreters

### 2018-11-10 paper/10_intro.tex
add some stuff

Futamura projections \cite{Futamura1999, futamura1971partial} reveal the close connection between 
compilers and interpreters. 
The first Futamura projection specifically shows that specializing an interpreter with respect to the input 
program yields an equivalent executable.
Partial evaluation \cite{DBLP:books/daglib/0072559} was the first proposed approach to Futamura projection,
which first identifies the binding-time of variables in the program, i.e., they can be known statically or dynamically, 
and then specializes the program with respect to the given input of program.
However, a precise binding-time analysis (BTA) is hard generally. As an alternative and practical approach to 
specialization and partial evaluation, multi-stage programming (MSP) \cite{taha1999multi, DBLP:conf/pepm/TahaS97} 
requires the stage annotations (i.e., binding-time annotations) from the programmers, these staging annotations 
control which part of the input program should be specialized.
As a simple example, we consider a power function: 
\begin{lstlisting}
def power(b: Int, e: Int): Int = 
  if (e == 0) 1 else e * power(b, e-1)
\end{lstlisting}
If \texttt{e} is statically known, say 5, then we can specialize \texttt{power} with respect to \texttt{e = 5} and 
generate a new specialized function as following:
\begin{lstlisting}
def power5(b: Int) = b * b * b * b * b
\end{lstlisting}

Abstract interpretation \cite{DBLP:conf/popl/CousotC77} as a semantic approach to construct sound static analysis 
by approximation is almost as old as Futamura projection.
Some recent progress such as Abstracting Abstract Machines uncovers a methodology to derive sound abstract interpreters 
from their concrete counterparts \cite{DBLP:journals/jfp/HornM12, DBLP:conf/icfp/HornM10, DBLP:journals/pacmpl/DaraisLNH17}.
\note{TR: Cousot was already systematic. Both Futamura and Cousot are results from the 1970s, but have
profited from more recent results (MSP and AAM) -- build a connection?} 
Given the structural similarity between concrete and abstract interpreters, intellectually 
it is natural to raise the question whether it is possible to stage an abstract interpreter, and how 
to effectively specialize an abstract interpreter having considered their functional dissimilarity. 
This paper studies the application of the first Futamura projection on abstract interpreters, 
and the approach to construct efficient abstract interpreters by multi-stage programming.

>>>>

Futamura projections \cite{Futamura1999, futamura1971partial} reveal the close connection between 
compilers and interpreters. 
The first Futamura projection specifically shows that specializing an interpreter with respect to the input 
program yields an equivalent executable.
Partial evaluation \cite{DBLP:books/daglib/0072559} was the first proposed approach to realize Futamura projections,
which first identifies the binding-time of variables in the program, i.e., they can be known statically or dynamically, 
and then specializes the program with respect to the given input of program.
However, precisely analyzing binding-time for an arbitray program is hard in general. As an alternative and more practical 
approach to specialization and partial evaluation, multi-stage programming (MSP) \cite{taha1999multi, DBLP:conf/pepm/TahaS97} 
requires the stage annotations (i.e., binding-time annotations) from the programmers, these staging annotations 
control which part of the input program should be specialized.
As a simple example, we consider a power function: 
\begin{lstlisting}
def power(b: Int, e: Int): Int = 
  if (e == 0) 1 else e * power(b, e-1)
\end{lstlisting}
If \texttt{e} is statically known, say 5, then we can specialize \texttt{power} with respect to \texttt{e = 5} and 
generate a new specialized function as following:
\begin{lstlisting}
def power5(b: Int) = b * b * b * b * b
\end{lstlisting}

Abstract interpretation \cite{DBLP:conf/popl/CousotC77} as a semantic approach to construct sound static analysis 
by approximation is almost as old as Futamura projection.
Some recent progress such as Abstracting Abstract Machines (AAM) uncovers a methodology to derive sound abstract interpreters 
from their concrete counterparts and has been applied to different variants of definitional interpreters and abstract machines
\cite{DBLP:journals/jfp/HornM12, DBLP:conf/icfp/HornM10, DBLP:journals/pacmpl/DaraisLNH17}.
\note{TR: Cousot was already systematic. Both Futamura and Cousot are results from the 1970s, but have
profited from more recent results (MSP and AAM) -- build a connection?} 
Given the structural similarity between concrete and abstract interpreters, intellectually 
it is natural to raise the question whether it is possible to stage an abstract interpreter, and how 
to effectively specialize an abstract interpreter having considered their functional dissimilarity. 
This paper studies the application of the first Futamura projection on abstract interpreters, 
and the approach to construct efficient abstract interpreters by multi-stage programming.

### 2018-11-10 paper/20_bg.tex
revise intro
add some stuff

\subsection{The Target Language} \label{bg_lang}

As higher-order functions had been widely introduced in mainstream programming languages \todo{cite}, in this paper, 
we target a tiny higher-order functional language \TLang, which is based on direct-style lambda-calculus
with the support of numbers, arithmetics, recursions and conditionals.
In section \ref{cases_imp}, we will see adding more imperative features to the language.
Since we are mostly interested in analyzing the dynamic behavior of the program, the static semantics and
type system will be omitted. We also assume all the input programs are well-typed.
The abstract syntax is described below:

\begin{lstlisting}
abstract class Expr
case class Lit(i: Int)
case class Var(x: String)
case class Lam(x: String, e: Expr)
case class App(e1: Expr, e2: Expr)
case class AOp(op: String, e1: Expr, e2: Expr)
case class Rec(x: String, e: Expr, body: Expr)
case class If0(cnd: Expr, thn: Expr, els: Expr)
\end{lstlisting}

We give the concrete semantics of \TLang by showing a big-step operational interpreter for it.
But before doing that, let's first see a general interface for interpreters, whether concrete or abstract, 
staged or unstaged. The trait \texttt{Semantics} first declares several abstract type members, such as \texttt{Ident},
\texttt{Addr} and etc., which will be instantiated concretely later. The abstract types \texttt{Env} and \texttt{Store}
are required to conform their specifications, \texttt{EnvSpec} and \texttt{StoreSpec}, which define 
the operations such as lookup and update on them, nevertheless the detail of them are omitted.  
The store component is necessary when modeling side effects such as mutations. 
The trait also requires the implementators provide the initial values of environment and store.

%TODO: talk about alloc here?

\begin{lstlisting}
trait Semantics {
  type R[+T]
  type Ident;    type Addr;   type Value;    type Ans
  type Env <: EnvSpec[R, Ident, Addr, Env]
  type Store <: StoreSpec[R, Addr, Value, Store]
  val $\rho$0: Env;     val $\sigma$0: Store
  def alloc(x: R[Ident], $\sigma$: Store): R[Addr]
  type EvalFun = (Expr, Env, Store) => Ans
  def eval(ev: EvalFun)(e: Expr, $\rho$: Env, $\sigma$: Store): Ans
  def fix(ev: EvalFun => EvalFun): EvalFun = 
    (e, $\rho$, $\sigma$) =>  ev(fix(ev))(e, $\rho$, $\sigma$)
  def eval_top(e: Expr): Ans = fix(eval)(e, $\rho$0, $\sigma$0)
}
\end{lstlisting}

The evaluator \texttt{eval} is a open-recursive function that takes one and three arguments: the first
one \texttt{ev} represents the recursive function itself; the rest three are the AST representing program 
will be executed, the environment that maps from identifiers to addresses, and the store that maps addresses to values.
This environment-and-store-passing style big-step interpreter also can be obtained by 
refunctionalizing \cite{DBLP:conf/ppdp/AgerBDM03, Wei:2018:RAA:3243631.3236800} a small-step CESK 
machine \cite{DBLP:conf/popl/FelleisenF87}.
We provide a function \texttt{fix} to close the recursion by providing the result of \texttt{fix(ev)} 
to argument \texttt{ev}. Finally, a default implementation for the top-level evaluation function \texttt{eval\_top} is defined.
As we will see later, the open-recursion style is beneficial when implementing abstract interpreter and 
staged abstract interpreter \todo{cite adi}.

Readers may notice an additional abstract type \texttt{R[+T]} is defined in the trait,
which means \texttt{R} represents a higher-kinded type and will used to specify the binding-time, 
i.e., staged or not. Since the LMS framework we are using allows us to specify binding-time purely
based on types, the unstaged higher-kinded type can be simply instantiated as an identity type constructor: 
\texttt{type NoRep[+T] = T}. In section \ref{stagedinterp}, \texttt{R} will be instantiated 
as a staging annotation. \todo{where talk about stage poly?}

Now we can instantiate a concrete interpreter. We first concretize the abstract types as follows.
Since the interpreter is concrete, the address space is infinite thus type \texttt{Int} is used for \texttt{Addr}.
The final returned type by the interpreter \texttt{Ans} is a pair of \texttt{Value} and \texttt{Store}, 
where the values are either closure or numbers.

\begin{lstlisting}
trait Concrete extends Semantics {
  type Ident = String;         type Addr = Int
  type Ans = (Value, Store);   abstract class Value
  case class CloV($\lambda$: Lam, $\rho$: Env) extends Value
  case class NumV(i: Int) extends Value
}
\end{lstlisting}

Then we define evaluation function that dispatches differently for different expression.
As we mentioned before, the type \texttt{R} is concretize as \texttt{NoRep}.
Note that an allocation function is used when we need to update the store: 
in the concrete case, \texttt{alloc} always returns a fresh address of the store.

\begin{lstlisting}
//TODO: maybe not show all the code
object NoRepConcreteEval extends Concrete {
  type R[+T] = NoRep[T]
  def alloc(x: Ident, $\sigma$: Store): Addr = $\sigma$.size + 1
  def eval(ev: EvalFun)(e: Expr, $\rho$: Env, $\sigma$: Store): Ans = 
    e match {
      case Lit(i) => (NumV(i), $\sigma$)
      case Var(x) => ($\sigma$($\rho$(x)), $\sigma$)
      case Lam(x, e) => (CloV(Lam(x,e), $\rho$), $\sigma$)
      case App(e1, e2) =>
        val (CloV(Lam(x,e), $\lambda\rho$), e1$\sigma$) = ev(e1, $\rho$, $\sigma$)
        val (e2v, e2$\sigma$) = ev(e2, $\rho$, e1$\sigma$)
        val $\alpha$ = alloc(x, e2$\sigma$)
        ev(e, $\lambda\rho$ + (x -> $\alpha$), e2$\sigma$ + ($\alpha$ -> e2v))
      case AOp(op, e1, e2) =>
        val (e1v: NumV, e1$\sigma$) = ev(e1, $\rho$, $\sigma$)
        val (e2v: NumV, e2$\sigma$) = ev(e2, $\rho$, e1$\sigma$)
        (evalArith(op, e1v, e2v), e2$\sigma$)
      case Rec(x, f, body) => 
        val $\alpha$ = alloc(x, $\sigma$);  val $\rho$_* = $\rho$ + (x -> $\sigma$)
        val (fv, f$\sigma$) = ev(f, $\rho$_*, $\sigma$)
        val $\sigma$_* = f$\sigma$ + ($\alpha$ -> fv);  ev(body, $\rho$_*, $\sigma$_*)
      case If0(cnd, thn, els) =>
        val (cndv: NumV, cnd$\sigma$) = ev(cnd, $\rho$, $\sigma$)
        if (cndv.i == 0) ev(thn, $\rho$, cnd$\sigma$)
        else ev(els, $\rho$, cnd$\sigma$)
    }
}
\end{lstlisting}

\subsection{From Interpreter to Abstract Interpreter}

\begin{lstlisting}
trait Abstract extends Semantics {
  type Ident = String;         case class Addr(x: Ident)
  type Ans = (Value, Store);   type Value = Set[AbsValue]
  abstract class AbsValue
  case class CloV($\lambda$: Lam, $\rho$: Env) extends AbsValue
  case class NumV() extends AbsValue
}
\end{lstlisting}

Address just record the variable name, more sensitivity in section ...

other big-step abstract interpreter, 
big-step abstract interpreter using delimited control operators \cite{Wei:2018:RAA:3243631.3236800},
monadic big-step abstract interpreter \cite{DBLP:journals/pacmpl/DaraisLNH17},
arrow-based abstract interpreter \cite{Keidel:2018:CSP:3243631.3236767}.

co-inductive caching/truncated depth-first evaluation \cite{DBLP:journals/pacmpl/DaraisLNH17, Wei:2018:RAA:3243631.3236800, Rosendahl:AbsIntPL}

\subsection{From Interpreter to Staged Interpreter} \label{stagedinterp}

\subsubsection{Multi-Stage Programming}
\begin{lstlisting}
val power5 = new DslDriver[Int, Int] {
  def power(b: Rep[Int], x: Int): Rep[Int] =
    if (x == 0) 1 else b * power(b, x-1)
  def snippet(x: Rep[Int]): Rep[Int] = power(x, 5)
}
\end{lstlisting}

Talk about power example. general idea; LMS framework in scala. MetaOCaml?

Other notable implementations of MSP exist in ML family, for examples, MetaML \cite{DBLP:conf/pepm/TahaS97} 
and MetaOCaml \cite{DBLP:conf/gpce/CalcagnoTHL03, DBLP:conf/flops/Kiselyov14}.
Compared with the LMS approach in Scala, MetaML/MetaOCaml use term-level annotations such as brackets, 
escape, and run. Notwithstanding, we use LMS in this paper, the idea of staging an abstract interpreter
still applies with other MSP implementations.

\subsubsection{Staged Interpreters}

>>>>

\input{21_bg_interp}
\input{23_bg_stage}
\input{22_bg_abs}

### 2018-11-10 paper/70_related.tex
fix paper
add some stuff

Abstracting Definitional Interpreter \cite{darais2017abstracting}

@@

>>>>

Abstracting Definitional Interpreter \cite{DBLP:journals/pacmpl/DaraisLNH17}.

@@

Other notable implementations of MSP exist in ML family, for examples, MetaML \cite{DBLP:conf/pepm/TahaS97} 
and MetaOCaml \cite{DBLP:conf/gpce/CalcagnoTHL03, DBLP:conf/flops/Kiselyov14}.
Compared with the LMS approach in Scala, MetaML/MetaOCaml use term-level annotations such as brackets, 
escape, and run. Notwithstanding, we use LMS in this paper, the idea of staging an abstract interpreter
still applies with other MSP implementations.
